Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-19 paper/paper.tex
fixes

This work presents the first Coq formalisation to our knowledge of
\emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
extraction of idiomatic OCaml code. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}). % As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

@@

Definition c_split : Coalg (TreeF unit int) (list int).
|{ x ~> match x with
        | nil => a_leaf tt
        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
                      a_node h l r
        end}|.
Defined.
\end{coqcode}
We prove that the coalgebra \coq{c_split} is recursive by showing
that it respects the ``less-than'' relation on the length of the lists.
%\begin{coqcode}
%Lemma split_fin : respects_relation c_split (@length int) lt.  (* *) Qed.
%
%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
%  := mk_wf_coalg wf_lt split_fin.
%\end{coqcode}
The code that we extract for \coq{hylo merge c_split} is the following:
\begin{ocamlcode}
let rec qsort = function
| [] -> [] | h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
  app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
code is similar to a hand-written \coq{qsort}.
The mergesort algorithm can be defined analogously and can be found in the
formalisation. % [REF LINK].

@@

Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
  (* [knapsack_alg wvs] as a respecful morphism *)
\end{coqcode}
The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
recursive coalgebra for \coq{nat}.
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
Coq's code extraction mechanism is unable to inline several definitions in this
case. We have manually inlined the extracted code for simplicity. The reader
can check in our \textbf{Artefact} that the extracted code can be trivially
inlined to produce the following:
\begin{ocamlcode}
let knapsack wvs x =
  let (y, _) =
    (let rec f x0 =
      if x0=0 then Uint63.of_int (0)
      else let fn := f (x0-1) in { lFix_out = { 
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape
  in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.

>>>>

This work presents the first Coq formalisation to our knowledge of
\emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
extraction of idiomatic OCaml code. The full mechanisation can be found on Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}).

% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

@@

Definition c_split : Coalg (TreeF unit int) (list int).
|{ x ~> match x with
        | nil => a_leaf tt
        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
                      a_node h l r
        end}|.
Defined.
\end{coqcode}
We prove that the coalgebra \coq{c_split} is recursive by showing
that it respects the ``less-than'' relation on the length of the lists.
%\begin{coqcode}
%Lemma split_fin : respects_relation c_split (@length int) lt.  (* *) Qed.
%
%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
%  := mk_wf_coalg wf_lt split_fin.
%\end{coqcode}
The code that we extract for \coq{hylo merge c_split} is the following:
\begin{ocamlcode}
let rec qsort = function
| [] -> [] | h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
  app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
code is similar to a hand-written \coq{qsort}.
The mergesort algorithm can be defined analogously and can be found in the
formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].

@@

Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
  (* [knapsack_alg wvs] as a respecful morphism *)
\end{coqcode}
The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
recursive coalgebra for \coq{nat}.
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
Coq's code extraction mechanism is unable to inline several definitions in this
case. We have manually inlined the extracted code for simplicity. The reader
can check in our artefact that the extracted code can be trivially
inlined to produce the following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) =
    (let rec f x0 =
      if x0=0 then Uint63.of_int (0)
      else let fn := f (x0-1) in { lFix_out = { 
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.
